---
title: "data and R packages"
author: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```
<style> #slideSty80 > p { margin-top: -80px; } </style>

## R data
- R comes with several inbuilt datasets

```{r, eval=TRUE}
data("iris")
head(iris)
```

# packages

## Packages
- R comes with basic functions that you will use on a daily basis
- Sometime you will required spelize functions, written by others
- R package is collection of functions and data written for a specific task

## Install a package
- To install an R package use the command **install.packages("package name")**
- In this example we install the package **beeswarm**

```{r, eval=FALSE}
install.packages("beeswarm")
```

## Load a package {#slideSty80}
- Now we load the package and use a function from it
```{r}
library("beeswarm")
data("iris")
beeswarm(iris[, c("Sepal.Length", "Sepal.Width")],
         pch = 16, col=c("red", "blue"),
         labels = c("Sepal Length", "Sepal Width"))
```

## Finding Help
- All functions in R are documented 
- **?** can be used to find documents
```{r}
?beeswarm
```

- **??** can be used to find something anywhere in R documents
```{r}
??beeswarm
```

# Now find more about the function **mean**

## Finding R package
- Google the problem
- Look into CRAN website www.cran.r-project.org/web/packages/
- Look into Bioconductor repo www.bioconductor.org
- **Note :** packages at Bioconductor uses different command to install that is:

```{r, eval=FALSE}
BiocManager::install("package name")
```

