---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# R data-types and structures

## Variable

A named container or storage that is used to reference a value in the code.

- This value can be of many types:
    + numeric: ```1, 2, 3.4```
    + character: ```"a", "book", "c"```
    + logical: ```TRUE, FALSE``` or simply ```T, F```
    + Integers, complex, and factors
    
## Variable

* Constraints on names:
    + must starts with a letter or a dot not followed by a number.
    + can only contain letters, numbers, underscore and the dot character "."
    
    ``` var1, .var1, var.name, v1 ```
    
    ``` 1v, .1v, var-1, var#1 ``` 

## Variable

```<-``` and ```=``` are used for assignment of values in R 

```{r}
  x <- 2
  y <- 3
  x+y
```

* Note that R variables are dynamically typed
```{r}
  x <- 2
  x <- "hello"
```


## Vector
A `vector` is the simplest data structure in R. it represents a container for a set of elements of the same data type. 
-  Examples:
```{r}
  c(1,3,5)
  length(c("test"))
  class(c(TRUE,TRUE,FALSE))
```

## Vector
`c` function is used to concatenate vectors.
```{r}
  x <- c(1,3,5)
  y <- c(2,4,6)
  z <- c(x,y)
  z
```

Remember: vectors have a single data type
```{r}
  x <- c(1,3,5)
  y <- c("a","b")
  c(x,y)
```

## Vector
Arithmetics are performed on vectors member-by-member
```{r}
  x <- c(1:3) # R way to create a sequence of numbers
  x
  x + 1
```

## Vector
and this extends to operations between vectors
```{r}
  x <- c(1,3,5)
  y <- c(2,4,6)
  x+y
```

## Vector
What will happen if the two vectors are of different lengths?
```{r, warning=FALSE}
  x <- c(1,3,5)
  y <- c(2,4,6,8,10)
  z <- x+y
```


## Vector
What will happen if the two vectors are of different lengths?
```{r, warning=FALSE}
  x <- c(1,3,5)
  y <- c(2,4,6,8,10)
  z <- x+y
  z
```

## Vector
Exercise: Create a vector of Logical values, assign it to a variable. Create a vector of Integers and assign it to a variable. 

Try subtracting the two vectors. What happened?

## Vector
Exercise: Create a vector of Logical values, assign it to a variable. Create a vector of Integers and assign it to a variable. 

Try subtracting the two vectors. What happened?
```{r}
x <- c(T,F,T)
y <- c(3,4,5)
x-y
```

logical values in R can be coerced to integers, i.e. `1` for `TRUE` and `0` for `FALSE`

## Vector
Retrieving elements from a vector is done by `indexing`

-  There are many ways to use indexing and the simplest is `numeric indexing`, i.e. a number that represents the position of an element in a vector.

-  `[]` is used for indexing. In R, indexing starts from `1`

```{r}
  x <- c(1,3,5,7,9)
  x[2]
  x[c(3,5)]
```

## Vector - Indexing
`indexing` can be used also for changing values in a vector

```{r}
  x <- c(1,3,5,7,9)
  x[2] <- 4 
  x
  x[c(3,5)] <- c(6,10)
  x
  x[-1] # negative indexing is used to remove elements
```

## Vector - Indexing
Logical values can be used for `indexing` too.

```{r}
  x <- c(1,3,5,7,9)
  x[c(T,F,T,F,T)]
  x[c(F,F,T,F,T)] <- c(6,10)
  x
  y <- x[c(T,F)] # what will happen??
```

## Vector - Indexing
Logical values can be used for `indexing` too.

```{r}
  x <- c(1,3,5,7,9)
  y <- x[c(T,F)] # what will happen??
  y
```

## Named Vectors 

R allows us to give names to each element in a vector, and use them to index:
```{r}
x <- c(2,4,6,8,10)
names(x) <- c("A", "B", "C", "D", "E")
x
x[c("A", "D")]
```

## Matrix
A `matrix` is a collection of data elements arranged in a two-dimensional rectangular layout
```{r}
x = matrix( 
   c(2, 4, 3, 1, 5, 7), # the data elements 
   nrow=2,              # number of rows 
   ncol=3)             # number of columns 
x
dim(x)  # return the dimensions of a matrix; #rows and #columns 
x[2,3] # note the ',' to indiacte row and column indexing
```

## Matrix - Names

Matrix can have row and/or column names:

```{r}
x <- matrix(data=1:10, nrow=2, ncol=5)
rownames(x) <- c("A", "B")
x
x["B", c(2,5)]
x["A",]
```



## Dataframe
A `dataframe` is used for storing data tables. It represents a list of vectors of equal length and can store vectors of different types.

```{r}
patientsID = c("p1", "p2", "p3") 
age = c(25, 35, 55) 
alive = c(TRUE, FALSE, NA) 
df = data.frame(patientsID, age, alive)       # df is a data frame
df
```

## Dataframe

$ is another way to index columns in a named data.frame

```{r}
df[,"alive"]
df$alive
```

## Dataframe

```{r}
patientsID = c("p1", "p2", "p3") 
age = c(25, 35, 55) 
alive = c(TRUE, FALSE, NA) 
df = data.frame(patientsID, age, alive)       # df is a data frame
df$patientsID  # factor type, we will talk about it later
df = data.frame(patientsID, age, alive, stringsAsFactors = F) 
df$patientsID
```


## List
A `list` is a generic vector that can contain multiple data types. Think of it as a vector of boxes, you can put anything in the box.
```{r}
x <- list(1, "a", c(TRUE, FALSE))
x
```


## List - indexing

One square bracket `[]` will retrieve the box, while double square brackets would unpack the box

```{r}
patientsID = c("p1", "p2", "p3", "p4", "p5") 
age = c(25, 35, 55) 
alive = c(TRUE, FALSE, TRUE, NA, FALSE) 
x <- list(patientsID,age,alive)
x[2]
x[[2]]
```


## List - indexing

```{r}
patientsID = c("p1", "p2", "p3", "p4", "p5") 
age = c(25, 35, 55) 
x <- list(patientsID,age)
x[[2]][3]
```

## List - indexing

Named lists provide another way to index them using the operator `$`

```{r}
patientsID = c("p1", "p2", "p3", "p4", "p5") 
age = c(25, 35, 55) 
x <- list("sampleID"=patientsID,"age"=age)
x$age
x[["sampleID"]]
```

## R data-types and structures

These are the main data types and structures in R.

We still need to talk about factors!

# R operations

## R Arithmetic operators

```{r, eval=FALSE}
x <- 5
y <- 3
+ x       # 5
- x       # -5
x + y     # 8
x - y     # 2
x * y     # 15
x / y     # 1.666667
x ^ y     # 125 , x**y is the same 
x %% y    # 2
x %/% y   # 1
```

## R Relational operators
```{r, eval=FALSE}
x <- 5
y <- 3
x < y    # FALSE
x > y    # TRUE
x <= y   # FALSE
x >= y   # TRUE
x == y   # FALSE
x != y   # TRUE

```


## R Logical operators

```{r, eval=FALSE}
x <- TRUE
y <- FALSE
!x       # FALSE  -- NOT
x && y   # FALSE  -- Logical AND
x || y   # TRUE   -- Logical OR
```


```{r}
x <- c(TRUE,TRUE,FALSE)
y <- c(TRUE,FALSE,FALSE)
x & y   # element-wise AND
x | y   # element-wise OR
```

# R control flow structures

## if-else 

This is a conditional flow structure by which the following steps of the program are dependent on the outcome of the condition.

```{r}
if(3>2){
  print("Yes") 
}
```

```{r}
if(3<2){
  print("Yes") 
}else{
  print("No")
}
```

## if-else 

```{r}
if(3<2){
  print("No") 
}else if(3>2){
  print("Yes")
}else{
  print("Maybe")
}
```

## if-else 

You can make the condition more complex
```{r}
x <- 5; y <- "food" 
if(x<2 && y == "drink"){
  print("Yes") 
}else{
  print("No")
}
```

## Loops

Loops are used to repeat a block of code until a condition is satisfied 

There are two flow structures for loops, i.e., `for` loops and `while` loops


## for loops

The most common use is to iterate through the elements of a vector

```{r}
myVec <- c("A", "B", "C")
for(letter in myVec){
  print(letter)
}
```

## for loops

One can also iterate through a vector using its indices

```{r}
myVec <- c(2, 5, 21)
for(i in 1:length(myVec)){
  print(myVec[i] + 3)
}
```

Note: `seq_along(myVec)` is a built-in function that might be a better option than using `1:length(myVec)`. The latter could break when `length(myVec)` is accidentally 0.

## while loops

Repeats executing the code enclosed while a condition is satisfied

```{r}
myVec <- c("a","b","c")
i <- 1
while(i <= length(myVec)){
  print(myVec[i])
  i <- i + 1
}
```

## Exercise

Create a R object from these vectors and return patients who are females and under the age of 30
```{r}
patientsID <- c("p1", "p2", "p3", "p4", "p5")
age <- c(15, 32, 44, 25, 39)
gender <- c("F","M","M","F","F")
```

## Exercise

Create a R object from these vectors and return patients who are females and under the age of 30
```{r}
patientsID <- c("p1", "p2", "p3", "p4", "p5")
age <- c(15, 32, 44, 25, 39)
gender <- c("F","M","M","F","F")

df <- data.frame(patientsID,age,gender,stringsAsFactors = F)
df[df$gender=="F" & df$age < 30,]
```

## Exercise

Create a R object from these vectors and change vector `insure` to `T` for each patient who is a female and above the age of 20
```{r}
patientsID <- c("p1", "p2", "p3", "p4", "p5")
age <- c(15, 32, 44, 25, 39)
gender <- c("F","M","M","F","F")
insure <- c(F,F,F,F,F)
```

## Exercise

Create a R object from these vectors and change vector `insure` to `T` for each patient who is a female and above the age of 20
```{r}
df <- data.frame(patientsID,age,gender,insure,stringsAsFactors = F)

for(patientInd in 1:dim(df)[1]){
  if(df[patientInd,"gender"]=="F" & df[patientInd,"age"]>20){
    df[patientInd,"insure"] = T
  }
}

df[,"insure"]

```

# functions

## functions

R does everything by copying. What do we mean by that?

```{r, results="hide"}
x <- 3
y <- x
x <- 2
x
y
```


## functions

R does everything by copying. What do we mean by that?

```{r}
x <- 3
y <- x
x <- 2
x
y
```


## functions

-  A `function` is a block of code that can be used multiple times. It is a way to structure the code.

-  R provides many built-in functions that ease writing programs and executing codes.

-  Examples: `print`, `:`, `c`, `in`

## functions

R also provides users with the ability to write their own custom functions

```{r}
printThanks <- function(){
  print("Thank you very much for running this code")
}

printThanks()
```

## functions

```{r}
isEven <- function(x){
  if(x%%2==0){
    return(TRUE)
  }
  return(FALSE)
}
isEven(4)
x <- isEven(3)
x
```

## functions

Variables defined inside a function are local to that function.

```{r, results="hide"}
x <- 5
testScope <- function(x){
  return(x+1)
}
testScope(4)
testScope(x)
x
```

## functions

Variables defined inside a function are local to that function.

```{r}
x <- 5
testScope <- function(x){
  return(x+1)
}
testScope(4)
testScope(x)
x
```



## *apply

Functions in R become super useful together with the built-in apply functions. *apply helps you apply a function to each element of a vector or row/column of a matrix. 


There are 3 main apply functions: `lapply`, `sapply` and `apply`. `lapply` - l stands for list, will return a list. `sapply` s stands for simple, will try to simplify the list. `apply` is `sapply` for matrices and data.frames. 

## sapply

Here is an sapply example:
```{r}
isEven <- function(x){
  if(x%%2==0){
    return(TRUE)
  }
  return(FALSE)
}

myVec <- c(4,3,2,1)
sapply(myVec, isEven)
```

## apply

`apply`, unlike sapply and lapply, takes 3 arguments: the data to apply over, which dimensions to apply over, and the function to apply. 

Heres a simple example, using the built-in `sum` function, which returns the sum of a vector:
```{r, results="hold"}
my.mat <- matrix(1:4, nrow=2, ncol=2)
my.mat
apply(my.mat, 1, sum)
apply(my.mat, 2, sum)
```

## functions
R has many built-in functions that can ease writing programs and you can find their syntax by using `?` operator

Examples:
```{r}
?mean
?rownames
?read.csv
```



## Exercise
Write a function will return T if a given element exists in a given vector.


## Exercise

```{r}
exists <- function(vector, element){
  
  for (v1 in vector) {
    if(v1 == element){
      return(TRUE)
    }
  }
  return(FALSE)
}
```


# Plots

## Plots {#slideSty40}
Plotting in R is easy:

```{r}
x = 1:10
y = x**2
plot(x, y)
```

## Plot {#slideSty40}
Lets change color

```{r}
plot(x, y, col = "red")
```

## Plot {#slideSty40}
Lets change point type

```{r}
plot(x, y, pch = 19)
```

## PCH 

PCH (plotting character) is an argument to specify point shapes

```{r, echo=FALSE, message=FALSE}
d = data.frame(p = c(0:25))
p <- ggplot() + scale_y_continuous(name = "") + scale_x_continuous(name = "")
p <- p + scale_shape_identity()
p <- p + geom_point(data = d, mapping = aes(x = p%%6, y = p%/%6, shape = p), 
                    size = 5, fill = "#91bfdb", color="black")

p <- p + geom_text(data = d, mapping = aes(x = p%%6, y = p%/%6 + 0.25, label = p),
                   size = 3) + scale_y_reverse()
p+theme_void()
```

## PCH 

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{font-family:Arial, sans-serif;font-size:14px;padding:10px 5px;border-style:solid;border-width:1px;overflow:hidden;word-break:normal;border-color:black;}
.tg th{font-family:Arial, sans-serif;font-size:14px;font-weight:normal;padding:10px 5px;border-style:solid;border-width:1px;overflow:hidden;word-break:normal;border-color:black;}
.tg .tg-4c4e{color:#3531ff;border-color:inherit;text-align:left;vertical-align:top}
</style>
<table class="tg">
  <tr>
    <td class="tg-4c4e">pch = 0, square</td>
    <td class="tg-4c4e">pch = 13, circle cross</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 1, circle</td>
    <td class="tg-4c4e">pch = 14, square and triangle down</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 2, triangle point up</td>
    <td class="tg-4c4e">pch = 15, filled square</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 3, plus</td>
    <td class="tg-4c4e">pch = 16, filled circle</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 4, cross</td>
    <td class="tg-4c4e">pch = 17, filled triangle point-up</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 5, diamond</td>
    <td class="tg-4c4e">pch = 18, filled diamond</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 6, triangle point down</td>
    <td class="tg-4c4e">pch = 19, solid circle</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 7, square cross</td>
    <td class="tg-4c4e">pch = 20, bullet (smaller circle)</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 8, star</td>
    <td class="tg-4c4e">pch = 21, filled circle blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 9, diamond plus</td>
    <td class="tg-4c4e">pch = 22, filled square blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 10, circle plus</td>
    <td class="tg-4c4e">pch = 23, filled diamond blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 11, triangles up and down</td>
    <td class="tg-4c4e">pch = 24, filled triangle point-up blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 12, square plus</td>
    <td class="tg-4c4e">pch = 25, filled triangle point down blue</td>
  </tr>
</table>


## Plot {#slideSty40}
- For PCH 21 to 25, background color can be specified

```{r}
plot(x, y, pch = 24, col="red", bg="green")
```


## Task

- plot data points in your favorite color and shape

## Plot Line {#slideSty40}

The option `"type"` can change the plot type

```{r}
plot(x, y, type = "l")
```


## Plot type

<table class="tg">
  <tr>
    <td class="tg-4c4e">"p"</td>
    <td class="tg-4c4e">for points</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"l"</td>
    <td class="tg-4c4e">for lines</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"b"</td>
    <td class="tg-4c4e">for both</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"c"</td>
    <td class="tg-4c4e">for the lines part alone of "b"</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"o"</td>
    <td class="tg-4c4e">for both 'overplotted'</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"h"</td>
    <td class="tg-4c4e">for 'histogram' like (or 'high-density') vertical lines</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"s"</td>
    <td class="tg-4c4e">for stair steps</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"S"</td>
    <td class="tg-4c4e">for other steps</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"n"</td>
    <td class="tg-4c4e">for no plotting</td>
  </tr>
</table>


## Plot {#slideSty40}
Lets plot both point and line

```{r}
plot(x, y, type = "o")
```

## Change line type {#slideSty40}
The option `"lty"` can change the line type

```{r}
plot(x, y, type = "o", lty=2)
```


## Line types in R
```{r, echo=FALSE, message=FALSE}
par(mar=c(0,0,0,0))
# Set up the plotting area
plot(NA, xlim=c(0,1), ylim=c(6.5, -0.5),
    xaxt="n", yaxt="n",
    xlab=NA, ylab=NA )
# Draw the lines
for (i in 0:6) {
    points(c(0.25,1), c(i,i), lty=i, lwd=2, type="l")
}
# Add labels
text(0, 0, "0. 'blank'"   ,  adj=c(0,.5))
text(0, 1, "1. 'solid'"   ,  adj=c(0,.5))
text(0, 2, "2. 'dashed'"  ,  adj=c(0,.5))
text(0, 3, "3. 'dotted'"  ,  adj=c(0,.5))
text(0, 4, "4. 'dotdash'" ,  adj=c(0,.5))
text(0, 5, "5. 'longdash'",  adj=c(0,.5))
text(0, 6, "6. 'twodash'" ,  adj=c(0,.5))
###dev.off()
```


## Task

- plot data points and line in your favorite color and shape 


## Plot Title {#slideSty40}
Give plot a title

```{r}
plot(x, y, main = "my 1st plot")
```

## Axis Name {#slideSty40}
Change plot axis name

```{r}
plot(x, y, main = "my 1st plot", xlab = "number", ylab = "value")
```

## Adding more to a plot {#slideSty50}
```{r}
x = 1:10
y = x**2
z = y + 1
plot(x, y, pch=19, col="red")
```


## Adding more to a plot {#slideSty80}
```{r}, fig.height = 4.0,  fig.align = "center"}
x = 1:10 
y = x**2 
z = y + 10
plot(x, y, pch=19, col="red")
lines(x, y, col="red")
points(x, z, col="blue")
lines(x, z, col="blue")
```


## Task

- plot data points and line in your favorite color and shape 


## Fixing axis limits {#slideSty60}
- In R plots are static 
- Adding new points can't change the axis limits
```{r}
plot(x, y, pch=19, col="red")
lines(x, y, col="red")
```


## Fixing axis limits {#slideSty50}
To overcome this issue specify the axis limits in advance
```{r}
ylimit = range(c(y,z))
plot(x, y, col="red", pch=19, ylim = ylimit)
lines(x, y, col="red")
points(x, z, col="blue", pch=15)
lines(x, z, col="blue")
```

## Adding legend {#slideSty50}
```{r}
plot(x, y, col="red", pch=19, ylim = c(1,110))
lines(x, y, col="red")
points(x, z, col="blue", pch=15)
lines(x, z, col="blue")
legend(1,100,legend=c("y","z"), col=c("red","blue"), pch=c(19,15))
```


## Barplot
```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors")
```

# How do we change the bar color?

## Barplot
```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors", col="red")
```


# How do we change border color?

## Barplot
```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors", col="red", border = NA)
```


## Barplot (Stacked) {#slideSty70}
```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
data = rbind(raptors, gsw)
colnames(data) = c("Q1", "Q2", "Q3", "Q4")
barplot(data, col=c("red", "gold"))
```

## Barplot (Grouped)
```{r}
barplot(data, col=c("red", "gold"), beside=TRUE)
```

## Barplot (add legend){#slideSty70}
```{r, fig.width= 8.0, fig.height = 6.0, fig.align = "center"}
barplot(data, col=c("red","gold"), beside=TRUE, legend=rownames(data),
        args.legend = list(x="topright"))
```


# Boxplot


## Boxplot {#slideSty30}
Same as barplot but data should be in column:
```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
boxplot(raptors, gsw)
```


## Boxplot {#slideSty30}
Add names to the box:
```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
boxplot(raptors, gsw, names=c("Raptors","GSW"))
```


## Boxplot (color) {#slideSty30}
Here HTML color codes are used: 
```{r}
boxplot(raptors, gsw, names=c("Raptors", "GSW"), 
        col=c("#d7301f","#dfc27d"), 
        border=c("#7f0000", "#8c510a"), main="NBA Finals")
```


## Task

- Add one more boxplot with data:
```{r}
bucks = c(23, 28, 25, 32)
```


## Using formula {#slideSty50}
Here HTML color codes are used: 
```{r}
bucks = c(23, 28, 25, 32)
df=data.frame(score=c(raptors, gsw, bucks), 
              team= c(rep("Raptors", 4), rep("GSW", 4), rep("Bucks", 4)))
boxplot(score~team, data=df)
```

## Using formula {#slideSty50}
Here HTML color codes are used: 
```{r}
bucks = c(23, 28, 25, 32)
df=data.frame(score=c(raptors, gsw, bucks), 
              team=c(rep("Raptors",4), rep("GSW",4), rep("Bucks",4)))
boxplot(score~team, data=df)
```

## Task

- Add one more boxplot with data:
```{r}

```



## Saving Plots for Publication
- For publications use PDF or PNG format
- To save plot, open file just before plot function
- Don't forget to close the file using **dev.off()**
```{r, eval=FALSE}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
bucks = c(23, 28, 25, 32)

pdf("NBA_finals.pdf", width = 4, height = 5)

boxplot(raptors, gsw, bucks, names=c("Raptors", "GSW", "Bucks"), 
        col=c("#d7301f","#dfc27d", "#41ab5d"), 
        border=c("#7f0000", "#8c510a", "#00441b"), main="NBA Finals")

dev.off()
```


## Saving Plots for Publication
- In **PNG** format, specify resolution using **res** 
- **res** should be at least 300 for publications  and 600 for presentation
- Also, don't forget to specify **units**

```{r, eval=FALSE}
png("figure-1.png", width = 4, height = 5, units = 'in', res = 600)

boxplot(raptors, gsw, bucks, names=c("Raptors", "GSW", "Bucks"), 
        col=c("#d7301f","#dfc27d", "#41ab5d"), 
        border=c("#7f0000", "#8c510a", "#00441b"), main="NBA Finals")

dev.off()
```

## R data
- R comes with several inbuilt datasets

```{r, eval=TRUE}
data("iris")
head(iris)
```

# packages

## Packages
- R comes with basic functions that you will use on a daily basis
- Sometime you will require specialized functions, written by others
- R package is a collection of functions and data written for a specific task

## Install a package
- To install an R package use the command **install.packages("package name")**
- In this example, we install the package **beeswarm**

```{r, eval=FALSE}
install.packages("beeswarm")
```

## Load a package {#slideSty80}
- Now we load the package and use a function from it
```{r}
library("beeswarm")
data("iris")
beeswarm(iris[, c("Sepal.Length", "Sepal.Width")],
         pch = 16, col=c("red", "blue"),
         labels = c("Sepal Length", "Sepal Width"))
```

## Finding Help
- All functions in R are documented 
- **?** can be used to find documents
```{r}
?beeswarm
```

- **??** can be used to find something anywhere in R documents
```{r}
??beeswarm
```

# Now find more about the function **mean**

## Finding R package
- Google the problem
- Look into CRAN website www.cran.r-project.org/web/packages/
- Look into Bioconductor repo www.bioconductor.org
- ** Note:** packages at Bioconductor uses a different command to install that is:

```{r, eval=FALSE}
BiocManager::install("package name")
```

# reading and writing files

## Save data {#slideSty80}
- R can save data in different formats
- Data can be saved in R native format

```{r, eval=TRUE}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
df = cbind(raptors, gsw)
saveRDS(df, "my_data.rds")
```
- To read it use readRDS function
```{r, eval=TRUE}
mydata = readRDS("my_data.rds")
head(mydata)
```

## Save data {#slideSty80}
- R can save data in CSV formats
```{r, eval=TRUE}
write.csv(df,file="my_data.csv")
```
- To read it use read.csv function
```{r, eval=TRUE}
mydata2 = read.csv("my_data.csv", row.names = 1)
head(mydata2)
```

## Excel files
- R package **openxlsx** can read and write Excel files
- Task: Install the package **openxlsx** and load it

## Write excel files
- To install the package
```{r, eval=FALSE}
install.packages("openxlsx")
```

- To write the Excel file
```{r, eval=TRUE}
library(openxlsx)
write.xlsx(df, "myData.xlsx")
```
- Now open the **myData.xlsx** file, add one more row, save and close the file


## Read excel files
```{r, eval=TRUE}
myxl = read.xlsx("myData.xlsx")
head(myxl)
```


## Caution
- Large Excel files may take a long time and memory to read/write
- Use R native **RDS**  files to read and write the data during the analysis
- To share data, prefer **CSV** format then Excel
- Save only final results/data in Excel format
 


# R Biostat. and data analysis

## R Biostat. basics

Useful functions:

- sum(x)      --> Sums the elements in x
- prod(x)      --> Product of the elements in x
- max(x)      --> Maximum element in x
- min(x)      --> Minimum element in x
- range(x)  --> Range (min to max) of elements in x


## R Biostat. basics

Useful functions:

- mean(x)   --> Mean (average value) of elements in x.
- median(x) --> Median  (middle  value)  of  elements  in x
- var(x)    --> Variance of elements in x
- sd(x)        --> Standard deviation of element in x
- cor(x,y)    --> Correlation between x and y
- cov(x,y)    --> Covariance between x and y
- quantile(x,p)    --> The pth quantile of x


## R Biostat. basics

Example: patients data

```{r, results='hold'}
set.seed(21341)
patients.df <- data.frame(
  "ID"=paste0("p",1:100)
  ,"Gender"=factor(sample(c("F","M"),100,replace = T))
  ,"Stage"=factor(sample(c("I","II","III","IV"),100,replace = T))
  ,"Age"=rgamma(100,shape = 50)
  ,"TumorVolume"=rgamma(100,shape = 10)
  ,stringsAsFactors = F)

head(patients.df)


```

## R Biostat. basics

```{r}
quantile(patients.df$Age)
```

## R Biostat. basics

```{r}
hist(patients.df$Age)
```


## R Biostat. basics

```{r}
summary(patients.df$Stage)
table(patients.df$Stage)
```

## R Biostat. basics

```{r}
boxplot(Age ~ Stage, patients.df
  , ylab="Age",xlab="Stage",main="Patients data")
```


## R Biostat. basics

T-test and Wilcoxon rank sum test

- T-test measures the difference in mean between two independent populations [assumes normal distribution]
- Wilcoxon rank sum test measures the difference in the median between two independent populations [non-parametric]


## R Biostat. basics

```{r}
boxplot(TumorVolume ~ Gender, patients.df
  , ylab="TumorVolume",xlab="Gender",main="Patients data")
```

## R Biostat. basics

Let's use the Wilcoxon rank sum test to see if Stage I and Stage IV patients are significantly different in regards to Tumor Volume

```{r}
  wilcox.test(x = patients.df$TumorVolume[patients.df$Gender=="M"]
              ,y = patients.df$TumorVolume[patients.df$Gender=="F"]
              ,mu = 0,alternative = "two.sided",paired = F,T)
```


## R Biostat. basics
Let's check the correlation between Age and Tumor volume

```{r}
plot(patients.df$Age,patients.df$TumorVolume)
```


## R Biostat. basics

- Pearson product moment correlation: The Pearson correlation evaluates the linear relationship between two continuous variables. 
- Spearman rank-order correlation: The Spearman correlation evaluates the monotonic relationship between two continuous or ordinal variables.

## R Biostat. basics
Let's check the correlation between Age and Tumor volume

```{r}
cor.test(patients.df$Age,patients.df$TumorVolume)
```



# Project

## Project
1- Download the following file from: 
https://drive.google.com/drive/folders/1Jqa5KxKyfYqMMkk1IOwKihOZw-IeedxT

2- Find the if there is a significant difference between TNBC and nonTNBC samples based on these genes [ESR1, PGR, ERBB2] and show that in a plot.
3- Compute the pairwise correlation between all the samples and show that in a plot. 

Hint: Install and use package **corrplot**





