---
title: "Plot"
author: ""
date: '2018-10-14'
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Plots
## Plots

Plotting in R is easy:

```{r}
x = 1:10
y = x**2
plot(x, y)
```

## Plot 

Lets change color
```{r}
plot(x, y, col = "red")
```

## Plot 

Lets change point type
```{r}
plot(x, y, pch = 19)
```

## PCH 

PCH (plotting character) is argument to specify point shapes

```{r, echo=FALSE, message=FALSE}
d = data.frame(p = c(0:25))
p <- ggplot() + scale_y_continuous(name = "") + scale_x_continuous(name = "")
p <- p + scale_shape_identity()
p <- p + geom_point(data = d, mapping = aes(x = p%%6, y = p%/%6, shape = p), 
                    size = 5, fill = "#91bfdb", color="black")

p <- p + geom_text(data = d, mapping = aes(x = p%%6, y = p%/%6 + 0.25, label = p),
                   size = 3) + scale_y_reverse()
p+theme_void()
```

## PCH 

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{font-family:Arial, sans-serif;font-size:14px;padding:10px 5px;border-style:solid;border-width:1px;overflow:hidden;word-break:normal;border-color:black;}
.tg th{font-family:Arial, sans-serif;font-size:14px;font-weight:normal;padding:10px 5px;border-style:solid;border-width:1px;overflow:hidden;word-break:normal;border-color:black;}
.tg .tg-4c4e{color:#3531ff;border-color:inherit;text-align:left;vertical-align:top}
</style>
<table class="tg">
  <tr>
    <td class="tg-4c4e">pch = 0, square</td>
    <td class="tg-4c4e">pch = 13, circle cross</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 1, circle</td>
    <td class="tg-4c4e">pch = 14, square and triangle down</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 2, triangle point up</td>
    <td class="tg-4c4e">pch = 15, filled square</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 3, plus</td>
    <td class="tg-4c4e">pch = 16, filled circle</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 4, cross</td>
    <td class="tg-4c4e">pch = 17, filled triangle point-up</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 5, diamond</td>
    <td class="tg-4c4e">pch = 18, filled diamond</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 6, triangle point down</td>
    <td class="tg-4c4e">pch = 19, solid circle</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 7, square cross</td>
    <td class="tg-4c4e">pch = 20, bullet (smaller circle)</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 8, star</td>
    <td class="tg-4c4e">pch = 21, filled circle blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 9, diamond plus</td>
    <td class="tg-4c4e">pch = 22, filled square blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 10, circle plus</td>
    <td class="tg-4c4e">pch = 23, filled diamond blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 11, triangles up and down</td>
    <td class="tg-4c4e">pch = 24, filled triangle point-up blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 12, square plus</td>
    <td class="tg-4c4e">pch = 25, filled triangle point down blue</td>
  </tr>
</table>


## Plot 

For PCH 21 to 25, background color can be specified
```{r}
plot(x, y, pch = 24, col="red", bg="green")
```


## Task

- plot data points in your favoriout color and shape



## Plot Line

The option `"type"` can change the plot type
```{r}
plot(x, y, type = "l")
```


## Plot type

<table class="tg">
  <tr>
    <td class="tg-4c4e">"p"</td>
    <td class="tg-4c4e">for points</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"l"</td>
    <td class="tg-4c4e">for lines</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"b"</td>
    <td class="tg-4c4e">for both</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"c"</td>
    <td class="tg-4c4e">for the lines part alone of "b"</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"o"</td>
    <td class="tg-4c4e">for both 'overplotted'</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"h"</td>
    <td class="tg-4c4e">for 'histogram' like (or 'high-density') vertical lines</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"s"</td>
    <td class="tg-4c4e">for stair steps</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"S"</td>
    <td class="tg-4c4e">for other steps</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"n"</td>
    <td class="tg-4c4e">for no plotting</td>
  </tr>
</table>


## Plot 

Lets plot both point and line
```{r}
plot(x, y, type = "o")
```

## Change line type

The option `"lty"` can change the line type
```{r}
plot(x, y, type = "o", lty=2)
```


## Line types in R


```{r, echo=FALSE, message=FALSE}
par(mar=c(0,0,0,0))
# Set up the plotting area
plot(NA, xlim=c(0,1), ylim=c(6.5, -0.5),
    xaxt="n", yaxt="n",
    xlab=NA, ylab=NA )
# Draw the lines
for (i in 0:6) {
    points(c(0.25,1), c(i,i), lty=i, lwd=2, type="l")
}
# Add labels
text(0, 0, "0. 'blank'"   ,  adj=c(0,.5))
text(0, 1, "1. 'solid'"   ,  adj=c(0,.5))
text(0, 2, "2. 'dashed'"  ,  adj=c(0,.5))
text(0, 3, "3. 'dotted'"  ,  adj=c(0,.5))
text(0, 4, "4. 'dotdash'" ,  adj=c(0,.5))
text(0, 5, "5. 'longdash'",  adj=c(0,.5))
text(0, 6, "6. 'twodash'" ,  adj=c(0,.5))
###dev.off()
```


## Task

- plot data points and line in your favoriout color and shape 


## Plot Title

Give plot a title
```{r}
plot(x, y, main = "my 1st plot")
```

## Axis Name

Give plot a title
```{r}
plot(x, y, main = "my 1st plot", xlab = "number", ylab = "value")
```

## Adding more to a plot

```{r}
x = 1:10
y = x**2
z = y + 1
plot(x, y, pch=19, col="red")
```

## Adding more to a plot {#slideID1}
<style> #slideID1 > p { margin-top: -80px; } </style>
```{r, fig.height = 4.20,  fig.align = "center"}
x = 1:10 
y = x**2 
z = y + 10
plot(x, y, pch=19, col="red")
lines(x, y, col="red")
points(x, z, col="blue")
lines(x, z, col="blue")
```


## Task

- plot data points and line in your favoriout color and shape 


## Fixing axis limits {#slideID}
<style> #slideID > p { margin-top: -50px; } </style>

R plots are static and adding new points can't change the axis limits
```{r}
plot(x, y, pch=19, col="red")
lines(x, y, col="red")
```

## Fixing axis limits {#slideID3}
<style> #slideID3 > p { margin-top: -50px; } </style>
To overcome this issue specify the axis limits in advance
```{r}
ylimit = range(c(y,z))
plot(x, y, col="red", pch=19, ylim = ylimit)
lines(x, y, col="red")
points(x, z, col="blue", pch=15)
lines(x, z, col="blue")
```

## Adding legend {#slideID4}
<style> #slideID4 > p { margin-top: -50px; } </style>
```{r}
plot(x, y, col="red", pch=19, ylim = c(1,110))
lines(x, y, col="red")
points(x, z, col="blue", pch=15)
lines(x, z, col="blue")
legend(1,100,legend=c("y","z"), col=c("red","blue"), pch=c(19,15))
```


## Barplot

```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors")
```

# How do we change the bar color?

## Barplot
```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors", col="red")
```


# How do we change border color?

## Barplot
```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors", col="red", border = NA)
```


## Barplot (Stacked) {#slideID40}
<style> #slideID40 > p { margin-top: -70px; } </style>
```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
data = rbind(raptors, gsw)
colnames(data) = c("Q1", "Q2", "Q3", "Q4")
barplot(data, col=c("red", "gold"))
```

## Barplot (Grouped)
```{r}
barplot(data, col=c("red", "gold"), beside=TRUE)
```

## Barplot (add legend){#slideID5}
<style> #slideID5 > p { margin-top: -70px; } </style>
```{r, fig.width= 8.0, fig.height = 6.0, fig.align = "center"}
barplot(data, col=c("red","gold"), beside=TRUE, legend=rownames(data),
        args.legend = list(x="topright"))
```


# Boxplot


## Boxplot {#slideID50}
<style> #slideID50 > p { margin-top: -30px; } </style>
Same as barplot but data should be in column:

```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
boxplot(raptors, gsw)
```


## Boxplot {#slideID51}
<style> #slideID51 > p { margin-top: -20px; } </style>
Add names to the box:

```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
boxplot(raptors, gsw, names=c("Raptors","GSW"))
```


## Boxplot (color)
Here HTML color codes are used: 
<style> #slideID5 > p { margin-top: -70px; } </style>
```{r}
boxplot(raptors, gsw, names=c("Raptors", "GSW"), 
        col=c("#d7301f","#dfc27d"), 
        border=c("#7f0000", "#8c510a"), main="NBA Finals")
```


## Task

- Add one more boxplot with data:
```{r}
bucks = c(23, 28, 25, 32)
```


## Saving Plots for Publication
- For publications use PDF or PNG format
- To save plot, open file just before plot function
- Don't forget to close the file using **dev.off()**
```{r, eval=FALSE}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
bucks = c(23, 28, 25, 32)

pdf("NBA_finals.pdf", width = 4, height = 5)

boxplot(raptors, gsw, bucks, names=c("Raptors", "GSW", "Bucks"), 
        col=c("#d7301f","#dfc27d", "#41ab5d"), 
        border=c("#7f0000", "#8c510a", "#00441b"), main="NBA Finals")

dev.off()
```


## Saving Plots for Publication
- In **PNG** format, specify resolution using **res** 
- **res** should be atleast 300 for publications  and 600 for presenation
- Also don't forget to specify **units**

```{r, eval=FALSE}
png("figure-1.png", width = 4, height = 5, units = 'in', res = 600)

boxplot(raptors, gsw, bucks, names=c("Raptors", "GSW", "Bucks"), 
        col=c("#d7301f","#dfc27d", "#41ab5d"), 
        border=c("#7f0000", "#8c510a", "#00441b"), main="NBA Finals")

dev.off()
```
