---
title: "Plot"
author: ""
date: '2018-10-14'
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Plots
## Plots

Plotting in R is easy:

```{r}
x = 1:10
y = x**2
plot(x, y)
```


## Plot 

Give plot a title
```{r}
plot(x, y)
```


## Plot 

Lets change color
```{r}
plot(x, y, col = "red")
```

## Plot 

Lets change point type
```{r}
plot(x, y, pch = 19)
```

## PCH 

PCH (plotting character) is argument to specify point shapes

```{r, echo=FALSE, message=FALSE}
d = data.frame(p = c(0:25))
p <- ggplot() + scale_y_continuous(name = "") + scale_x_continuous(name = "")
p <- p + scale_shape_identity()
p <- p + geom_point(data = d, mapping = aes(x = p%%6, y = p%/%6, shape = p), 
                    size = 5, fill = "#91bfdb", color="black")

p <- p + geom_text(data = d, mapping = aes(x = p%%6, y = p%/%6 + 0.25, label = p),
                   size = 3) + scale_y_reverse()
p+theme_void()
```

## PCH 

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{font-family:Arial, sans-serif;font-size:14px;padding:10px 5px;border-style:solid;border-width:1px;overflow:hidden;word-break:normal;border-color:black;}
.tg th{font-family:Arial, sans-serif;font-size:14px;font-weight:normal;padding:10px 5px;border-style:solid;border-width:1px;overflow:hidden;word-break:normal;border-color:black;}
.tg .tg-4c4e{color:#3531ff;border-color:inherit;text-align:left;vertical-align:top}
</style>
<table class="tg">
  <tr>
    <td class="tg-4c4e">pch = 0, square</td>
    <td class="tg-4c4e">pch = 13, circle cross</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 1, circle</td>
    <td class="tg-4c4e">pch = 14, square and triangle down</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 2, triangle point up</td>
    <td class="tg-4c4e">pch = 15, filled square</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 3, plus</td>
    <td class="tg-4c4e">pch = 16, filled circle</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 4, cross</td>
    <td class="tg-4c4e">pch = 17, filled triangle point-up</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 5, diamond</td>
    <td class="tg-4c4e">pch = 18, filled diamond</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 6, triangle point down</td>
    <td class="tg-4c4e">pch = 19, solid circle</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 7, square cross</td>
    <td class="tg-4c4e">pch = 20, bullet (smaller circle)</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 8, star</td>
    <td class="tg-4c4e">pch = 21, filled circle blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 9, diamond plus</td>
    <td class="tg-4c4e">pch = 22, filled square blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 10, circle plus</td>
    <td class="tg-4c4e">pch = 23, filled diamond blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 11, triangles up and down</td>
    <td class="tg-4c4e">pch = 24, filled triangle point-up blue</td>
  </tr>
  <tr>
    <td class="tg-4c4e">pch = 12, square plus</td>
    <td class="tg-4c4e">pch = 25, filled triangle point down blue</td>
  </tr>
</table>


## Plot 

For PCH 21 to 25, background color can be specified
```{r}
plot(x, y, pch = 24, col="red", bg="green")
```


## Task

- plot data points in your favoriout color and shape



## Plot Line

The option `"type"` can change the plot type
```{r}
plot(x, y, type = "l")
```


## Plot type

<table class="tg">
  <tr>
    <td class="tg-4c4e">"p"</td>
    <td class="tg-4c4e">for points</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"l"</td>
    <td class="tg-4c4e">for lines</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"b"</td>
    <td class="tg-4c4e">for both</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"c"</td>
    <td class="tg-4c4e">for the lines part alone of "b"</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"o"</td>
    <td class="tg-4c4e">for both 'overplotted'</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"h"</td>
    <td class="tg-4c4e">for 'histogram' like (or 'high-density') vertical lines</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"s"</td>
    <td class="tg-4c4e">for stair steps</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"S"</td>
    <td class="tg-4c4e">for other steps</td>
  </tr>
  <tr>
    <td class="tg-4c4e">"n"</td>
    <td class="tg-4c4e">for no plotting</td>
  </tr>
</table>


## Plot 

Lets plot both point and line
```{r}
plot(x, y, type = "b")
```

## Task

- plot data points and line in your favoriout color and shape 


## Plot Title

Give plot a title
```{r}
plot(x, y, main = "my 1st plot")
```

## Axis Name

Give plot a title
```{r}
plot(x, y, main = "my 1st plot", xlab = "number", ylab = "value")
```


## Barplot

```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors")
```

# How do we change the bar color?

## Barplot

```{r}
raptors = c(33,27,26,28)
barplot(raptors, main="Toronto Raptors", col="red")
```

## Barplot

```{r}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
data = data.frame(raptors=raptors, gsw=gsw)
barplot(t(as.matrix(data)), col=c("red", "blue"), beside=TRUE)
```


R has three main basic data types:

Character:
```{r chars, echo=TRUE, results='hide'}
"a"
"Elephant"
"Hello World"
```

Numeric:
```{r nums, echo=TRUE, results='hide'}
1
34
```

Logical:
```{r logics, results='hide'}
TRUE
FALSE
```

## Other Data Types

Other Data Types you might meet include:

- Integers `1L`
- Complex numbers `1 + i`
- Factors

More on factors later.

## Variables

Assigning data to variables allows us to refer to data later on in our code. R has two main assignment operators, `=` and `<-`. 

`->` also exists but is frowned upon, `<-` is prefered. 

```{r}
a <- 2
b <- 3
a + b
```

We also met our first built in operator, `+`. 

R supports most of the normal math operators as you would expect. `?Arithmetic` will bring up the help page for arithmetic operators. 

## Variable Names

R is quite flexible in allowed variable names. Here are the rules:

- Starts with a letter or `.` not followed by a number
- Letters, numbers, dot or underline characters can be included
- Correct: “valid.name”, ”valid name”, ”valid2name3”
- Incorrect: ”valid name”, ”valid-name”, ”1valid2name3”

**Note: `.` has no special meaning in R**

## Vectors

Fundementally, almost everything in R (except more complex, custom objects defined by users or packages) is a Vector. Everything we met above was a vector of length 1:

```{r}
length(43)
length("Hello World")
length(TRUE)
```
