---
title: "Read and write data in R"
author: ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```
<style> #slideSty80 > p { margin-top: -80px; } </style>

## Save data {#slideSty80}
- R can save data in different formats
- Data can be saved in R native format

```{r, eval=TRUE}
raptors = c(33,27,26,28)
gsw = c(32, 25, 31, 22)
df = cbind(raptors, gsw)
saveRDS(df, "my_data.rds")
```
- To read it use readRDS function
```{r, eval=TRUE}
mydata = readRDS("my_data.rds")
head(mydata)
```

## Save data {#slideSty80}
- R can save data in CSV formats
```{r, eval=TRUE}
write.csv(df,file="my_data.csv")
```
- To read it use read.csv function
```{r, eval=TRUE}
mydata2 = read.csv("my_data.csv", row.names = 1)
head(mydata2)
```

## Excel files
- R package **openxlsx** can read and write Excel files
- Task: Install the package **openxlsx** and load it

## Write excel files
- To install the package
```{r, eval=FALSE}
install.packages("openxlsx")
```

- To write the Excel file
```{r, eval=TRUE}
library(openxlsx)
write.xlsx(df, "myData.xlsx")
```
- Now open the **myData.xlsx** file, add one more row, save and close the file


## Read excel files
```{r, eval=TRUE}
myxl = read.xlsx("myData.xlsx")
head(myxl)
```


## Caution
- Large Excel files may take long time and memory to read/write
- Use R native **RDS**  files to read and write the data during the analysis
- To share data, prefer **CSV** format then Excel
- Save only final results/data in Excel format
 

