---
title: "R for beginners Notebook"
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R Biostat. and data analysis

## R Biostat. basics

Useful functions:

- sum(x)	  --> Sums the elements in x
- prod(x)	  --> Product of the elements in x
- max(x)	  --> Maximum element in x
- min(x)	  --> Minimum element in x
- range(x)  --> Range (min to max) of elements in x


## R Biostat. basics

Useful functions:

- mean(x)   --> Mean (average value) of elements in x.
- median(x) --> Median  (middle  value)  of  elements  in x
- var(x)    --> Variance of elements in x
- sd(x)	    --> Standard deviation of element in x
- cor(x,y)	--> Correlation between x and y
- cov(x,y)	--> Covariance between x and y
- quantile(x,p)	--> The pth quantile of x


## R Biostat. basics

Example: patients data

```{r, results='hold'}
patients.df <- data.frame(
  "ID"=paste0("p",1:100)
  ,"Age"=sample(30:70,100,replace = T)
  ,"Stage"=factor(sample(c("I","II","III","IV"),100,replace = T))
  ,stringsAsFactors = F)

head(patients.df)


```

## R Biostat. basics

Example: patients data

```{r}

quantile(patients.df$Age)
hist(patients.df$Age)


```



## R Biostat. basics

Example: patients data

```{r, results='hold'}
summary(patients.df$Stage)

boxplot(Age ~ Stage, patients.df)

```











