---
title: "R for beginners Notebook"
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R data-types and structures

## Variable

A named container or storage that is used to reference a value in the code.

- This value can be of many types:
    + numeric: ```1, 2, 3.4```
    + character: ```"a", "book", "c"```
    + logical: ```TRUE, FALSE``` or simply ```T, F```
    + Integers, complex, and factors
    
## Variable

* Constraints on names:
    + must starts with a letter or a dot not followed by a number.
    + can only contain letters, numbers, underscore and the dot character "."
    
    ``` var1, .var1, var.name, v1 ```
    
    ``` 1v, .1v, var-1, var#1 ``` 

## Variable

```<-``` and ```=``` are used for assignment of values in R 

```{r}
  x <- 2
  y <- 3
  x+y
```

* Note that R variables are dynamically typed
```{r}
  x <- 2
  x <- "hello"
```


## Vector
A `vector` is the simplest data structure in R. it represents a container for a set of elements of the same data type. 
-  Examples:
```{r}
  c(1,3,5)
  length(c("test"))
  class(c(TRUE,TRUE,FALSE))
```

## Vector
`c` function is used to concatenate vectors.
```{r}
  x <- c(1,3,5)
  y <- c(2,4,6)
  z <- c(x,y)
  z
```

Remember: vectors have a single data type
```{r}
  x <- c(1,3,5)
  y <- c("a","b")
  c(x,y)
```

## Vector
Arithmetics are performed on vectors member-by-member
```{r}
  x <- c(1:3) # R way to create a sequence of numbers
  x
  x + 1
```

## Vector
and this extends to operations between vectors
```{r}
  x <- c(1,3,5)
  y <- c(2,4,6)
  x+y
```

## Vector
What will happen if the two vectors are of different lengths?
```{r, warning=FALSE}
  x <- c(1,3,5)
  y <- c(2,4,6,8,10)
  z <- x+y
```


## Vector
What will happen if the two vectors are of different lengths?
```{r, warning=FALSE}
  x <- c(1,3,5)
  y <- c(2,4,6,8,10)
  z <- x+y
  z
```

## Vector
Exercise: Create a vector of Logical values, assign it to a variable. Create a vector of Integers and assign it to a variable. 

Try subtracting the two vectors. What happened?

## Vector
Exercise: Create a vector of Logical values, assign it to a variable. Create a vector of Integers and assign it to a variable. 

Try subtracting the two vectors. What happened?
```{r}
x <- c(T,F,T)
y <- c(3,4,5)
x-y
```

logical values in R can be coerced to integers, i.e. `1` for `TRUE` and `0` for `FALSE`

## Vector
Retrieving elements from a vector is done by `indexing`

-  There are many ways to use indexing and the simplest is `numeric indexing`, i.e. a number that represents the position of an element in a vector.

-  `[]` is used for indexing. In R, indexing starts from `1`

```{r}
  x <- c(1,3,5,7,9)
  x[2]
  x[c(3,5)]
```

## Vector - Indexing
`indexing` can be used also for changing values in a vector

```{r}
  x <- c(1,3,5,7,9)
  x[2] <- 4 
  x
  x[c(3,5)] <- c(6,10)
  x
  x[-1] # negative indexing is used to remove elements
```

## Vector - Indexing
Logical values can be used for `indexing` too.

```{r}
  x <- c(1,3,5,7,9)
  x[c(T,F,T,F,T)]
  x[c(F,F,T,F,T)] <- c(6,10)
  x
  y <- x[c(T,F)] # what will happen??
```

## Vector - Indexing
Logical values can be used for `indexing` too.

```{r}
  x <- c(1,3,5,7,9)
  y <- x[c(T,F)] # what will happen??
  y
```

## Named Vectors 

R allows us to give names to each element in a vector, and use them to index:
```{r}
x <- c(2,4,6,8,10)
names(x) <- c("A", "B", "C", "D", "E")
x
x[c("A", "D")]
```

## Matrix
A `matrix` is a collection of data elements arranged in a two-dimensional rectangular layout
```{r}
x = matrix( 
   c(2, 4, 3, 1, 5, 7), # the data elements 
   nrow=2,              # number of rows 
   ncol=3)             # number of columns 
x
dim(x)  # return the dimensions of a matrix; #rows and #columns 
x[2,3] # note the ',' to indiacte row and column indexing
```

## Matrix - Names

Matrix can have row and/or column names:

```{r}
x <- matrix(data=1:10, nrow=2, ncol=5)
rownames(x) <- c("A", "B")
x
x["B", c(2,5)]
x["A",]
```



## Dataframe
A `dataframe` is used for storing data tables. It represents a list of vectors of equal length and can store vectors of different types.

```{r}
patientsID = c("p1", "p2", "p3") 
age = c(25, 35, 55) 
alive = c(TRUE, FALSE, NA) 
df = data.frame(patientsID, age, alive)       # df is a data frame
df
```

## Dataframe

$ is another way to index columns in a named data.frame

```{r}
df[,"alive"]
df$alive
```

## Dataframe

```{r}
patientsID = c("p1", "p2", "p3") 
age = c(25, 35, 55) 
alive = c(TRUE, FALSE, NA) 
df = data.frame(patientsID, age, alive)       # df is a data frame
df$patientsID  # factor type, we will talk about it later
df = data.frame(patientsID, age, alive, stringsAsFactors = F) 
df$patientsID
```


## List
A `list` is a generic vector that can contain multiple data types. Think of it as a vector of boxes, you can put anything in the box.
```{r}
x <- list(1, "a", c(TRUE, FALSE))
x
```


## List - indexing

One square bracket `[]` will retrieve the box, while double square brackets would unpack the box

```{r}
patientsID = c("p1", "p2", "p3", "p4", "p5") 
age = c(25, 35, 55) 
alive = c(TRUE, FALSE, TRUE, NA, FALSE) 
x <- list(patientsID,age,alive)
x[2]
x[[2]]
```


## List - indexing

```{r}
patientsID = c("p1", "p2", "p3", "p4", "p5") 
age = c(25, 35, 55) 
x <- list(patientsID,age)
x[[2]][3]
```

## List - indexing

Named lists provide another way to index them using the operator `$`

```{r}
patientsID = c("p1", "p2", "p3", "p4", "p5") 
age = c(25, 35, 55) 
x <- list("sampleID"=patientsID,"age"=age)
x$age
x[["sampleID"]]
```

## R data-types and structures

These are the main data types and structures in R.

We still need to talk about factors!

# R operations

## R Arithmetic operators

```{r, echo=TRUE, results='hide'}
x <- 5
y <- 3
+ x       # 5
- x       # -5
x + y     # 8
x - y     # 2
x * y     # 15
x / y     # 1.666667
x ^ y     # 125 , x**y is the same 
x %% y    # 2
x %/% y   # 1
```

## R Relational operators
```{r, echo=TRUE, results='hide'}
x <- 5
y <- 3
x < y    # FALSE
x > y    # TRUE
x <= y   # FALSE
x >= y   # TRUE
x == y   # FALSE
x != y   # TRUE

```


## R Logical operators

```{r, echo=TRUE, results='hide'}
x <- TRUE
y <- FALSE
!x       # FALSE  -- NOT
x && y   # FALSE  -- Logical AND
x || y   # TRUE   -- Logical OR
```


```{r}
x <- c(TRUE,TRUE,FALSE)
y <- c(TRUE,FALSE,FALSE)
x & y   # element-wise AND
x | y   # element-wise OR
```

# R control flow structures

## if-else 

This is a conditional flow structure by which the following steps of the program are dependent on the outcome of the condition.

```{r}
if(3>2){
  print("Yes") 
}
```

```{r}
if(3<2){
  print("Yes") 
}else{
  print("No")
}
```

## if-else 

```{r}
if(3<2){
  print("No") 
}else if(3>2){
  print("Yes")
}else{
  print("Maybe")
}
```

## if-else 

You can make the condition more complex
```{r}
x <- 5; y <- "food" 
if(x<2 && y == "drink"){
  print("Yes") 
}else{
  print("No")
}
```

## Loops

Loops are used to repeat a block of code until a condition is satisfied 

There are two flow structures for loops, i.e., `for` loops and `while` loops


## for loops

The most common use is to iterate through the elements of a vector

```{r}
myVec <- c("A", "B", "C")
for(letter in myVec){
  print(letter)
}
```

## for loops

One can also iterate through a vector using its indices

```{r}
myVec <- c(2, 5, 21)
for(i in 1:length(myVec)){
  print(myVec[i] + 3)
}
```

Note: `seq_along(myVec)` is a built-in function that might be a better option than using `1:length(myVec)`. The latter could break when `length(myVec)` is accidentally 0.

## while loops

Repeats executing the code enclosed while a condition is satisfied

```{r}
myVec <- c("a","b","c")
i <- 1
while(i <= length(myVec)){
  print(myVec[i])
  i <- i + 1
}
```

## Exercise

Create a R object from these vectors and return patients who are females and under the age of 30
```{r}
patientsID <- c("p1", "p2", "p3", "p4", "p5")
age <- c(15, 32, 44, 25, 39)
gender <- c("F","M","M","F","F")
```

## Exercise

Create a R object from these vectors and return patients who are females and under the age of 30
```{r}
patientsID <- c("p1", "p2", "p3", "p4", "p5")
age <- c(15, 32, 44, 25, 39)
gender <- c("F","M","M","F","F")

df <- data.frame(patientsID,age,gender,stringsAsFactors = F)
df[df$gender=="F" & df$age < 30,]
```

## Exercise

Create a R object from these vectors and change vector `insure` to `T` for each patient who is a female and above the age of 20
```{r}
patientsID <- c("p1", "p2", "p3", "p4", "p5")
age <- c(15, 32, 44, 25, 39)
gender <- c("F","M","M","F","F")
insure <- c(F,F,F,F,F)
```

## Exercise

Create a R object from these vectors and change vector `insure` to `T` for each patient who is a female and above the age of 20
```{r}
df <- data.frame(patientsID,age,gender,insure,stringsAsFactors = F)

for(patientInd in 1:dim(df)[1]){
  if(df[patientInd,"gender"]=="F" & df[patientInd,"age"]>20){
    df[patientInd,"insure"] = T
  }
}

df[,"insure"]

```

# functions

## functions

R does everything by copying. What do we mean by that?

```{r, results="hide"}
x <- 3
y <- x
x <- 2
x
y
```


## functions

R does everything by copying. What do we mean by that?

```{r}
x <- 3
y <- x
x <- 2
x
y
```


## functions

-  A `function` is a block of code that can be used multiple times. It is a way to structure the code.

-  R provides many built-in functions that ease writing programs and executing codes.

-  Examples: `print`, `:`, `c`, `in`

## functions

R also provides users with the ability to write their own custom functions

```{r}
printThanks <- function(){
  print("Thank you very much for running this code")
}

printThanks()
```

## functions

```{r}
isEven <- function(x){
  if(x%%2==0){
    return(TRUE)
  }
  return(FALSE)
}
isEven(4)
x <- isEven(3)
x
```

## functions

Variables defined inside a function are local to that function.

```{r, results="hide"}
x <- 5
testScope <- function(x){
  return(x+1)
}
testScope(4)
testScope(x)
x
```

## functions

Variables defined inside a function are local to that function.

```{r}
x <- 5
testScope <- function(x){
  return(x+1)
}
testScope(4)
testScope(x)
x
```



## *apply

Functions in R become super useful together with the built-in apply functions. *apply helps you apply a function to each element of a vector or row/column of a matrix. 


There are 3 main apply functions: `lapply`, `sapply` and `apply`. `lapply` - l stands for list, will return a list. `sapply` s stands for simple, will try to simplify the list. `apply` is `sapply` for matrices and data.frames. 

## sapply

Here is an sapply example:
```{r}
isEven <- function(x){
  if(x%%2==0){
    return(TRUE)
  }
  return(FALSE)
}

myVec <- c(4,3,2,1)
sapply(myVec, isEven)
```

## apply

`apply`, unlike sapply and lapply, takes 3 arguments: the data to apply over, which dimensions to apply over, and the function to apply. 

Heres a simple example, using the built-in `sum` function, which returns the sum of a vector:
```{r, results="hold"}
my.mat <- matrix(1:4, nrow=2, ncol=2)
my.mat
apply(my.mat, 1, sum)
apply(my.mat, 2, sum)
```

## functions
R has many built-in functions that can ease writing programs and you can find their syntax by using `?` operator

Examples:
```{r}
?mean
?rownames
?read.csv
```



## Exercise
Write a function will return T if a given element exists in a given vector.


## Exercise

```{r}
exists <- function(vector, element){
  
  for (v1 in vector) {
    if(v1 == element){
      return(TRUE)
    }
  }
  return(FALSE)
}
```






