---
title: "R for beginners Notebook"
output:
  ioslides_presentation: default
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R packages

## R Packages

- R packages are collections of functions and data sets developed by the community.
- They increase the power of R by improving existing base R functionalities, or by adding new ones. 
- Typically, a package will include code (not only R code!), documentation for the package and the functions inside, some tests to check everything works as it should, and data sets.


## R Packages

There are many repositories that hosts R packages for public use:
- CRAN: the official R repository.
- Bioconductor: a repository intended for open source software for bioinformatics.
- others: github (mostly for under development packages).

## R Packages

Installing a package that is in CRAN is simple!

```{r,results='hide',warning=FALSE}
install.packages("packageName",repos = "http://cran.us.r-project.org")
```
Example:
```{r,warning=FALSE,results='hide'}
install.packages("RXKCD",repos = "http://cran.us.r-project.org")
```

## R Packages

To load a package:
```{r}
library(RXKCD)
getXKCD(386)
```


## R Packages

Bioconductor


